--- 
title: "Modelling Land-use Impacts on Biodiversity - Getting Started with PREDICTS" 
author: "Tim Newbold" 
output: html_document 
---

This document will introduce the PREDICTS database, show you how to get the data, and how to start with some simple models of land-use impacts.

# Package dependencies
* performance
* predictsFunctions
* StatisticalModels

The predictsFunctions and StatisticalModels packages can be installed from GitHub:

```{r, echo=TRUE, warning=FALSE, message=FALSE,results=FALSE,eval=FALSE}
install.packages("performance")
library(remotes)
remotes::install_github("timnewbold/predicts-demo",subdir = "predictsFunctions")
remotes::install_github("timnewbold/StatisticalModels")
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, results=FALSE}
library(performance)
library(predictsFunctions)
library(StatisticalModels)
print(sessionInfo())
```

# The PREDICTS Database

## Getting the PREDICTS Database

You can get started quickly using the code below. If you intend to use the PREDICTS database in your work, please download it from the [Natural History Museum data portal](https://data.nhm.ac.uk/dataset/the-2016-release-of-the-predicts-database), to give credit for your use.

```{r, echo=TRUE}
# Download and import the PREDICTS data
myFile <- url("https://www.dropbox.com/s/pb1mdiel8o22186/database.rds?dl=1")
predicts <- readRDS(myFile)
print(head(predicts))
```

## Getting the Database Ready for Analysis

There are some standard steps to take before working with the PREDICTS database:
1. Correcting abundance values where sampling effort varies among the sites sampled in a single study.
2. Merging sites with the same coordinates.

```{r, echo=TRUE, message=FALSE, results=FALSE}
# Correct effort-sensitive abundance measures (assumes linear relationship between effort and recorded abundance)
predicts <- predictsFunctions::CorrectSamplingEffort(diversity = predicts)

# Merge sites that have the same coordinates (e.g. multiple traps on a single transect)
predicts <- predictsFunctions::MergeSites(diversity = predicts)
```

A lot of models of the PREDICTS database are of assemblage-level biodiversity measures. You can calculate these site-level measures using the *SiteMetrics* function.

```{r, echo=TRUE}
# Calculate site metrics of diversity - currently only species richness and total abundance
sites <- SiteMetrics(diversity = predicts,
                     extra.cols = c("Predominant_land_use",
                                    "SSB","SSBS","Biome"))
```

# Modelling the Impacts of Land Use

## Site-level biodiversity

We will start with some simple models of site-level biodiversity as a function only of land use.

### Data preparation

First, we will rearrange the land-use classification a bit.

```{r, echo=TRUE}
sites$LandUse <- paste(sites$Predominant_land_use)
# Drop classification where land use could not be identified
sites$LandUse[(sites$LandUse=="Cannot decide")] <- NA
# Drop classification where the stage of recovery of secondary vegetation is unknown
sites$LandUse[(sites$LandUse=="Secondary vegetation (indeterminate age)")] <- NA
# Now make the variable a factor, and set the reference level to primary vegetation
sites$LandUse <- factor(sites$LandUse)
sites$LandUse <- relevel(sites$LandUse,ref="Primary vegetation")
```

### Selecting random effects

We have to model the PREDICTS data using mixed-effects models, to account for the hierarchical structure of the database. We will start by investigating different possible random-effects structures for our models. All PREDICTS models need to have a random intercept of Study identity, since biodiversity measures are not comparable between studies.

We will compare here a model with a random intercept only of study, with one that contains a nested random intercept representing spatial blocks of sites within studies (to account for the spatial structuring of sites).

```{r, echo=TRUE}
# Fit a model with a random intercept only of study identity
r1 <- GLMER(modelData = sites,responseVar = "Species_richness",
            fitFamily = "poisson",fixedStruct = "LandUse",
            randomStruct = "(1|SS)",REML = FALSE)

# Try a model that also has a nested random intercept for spatial block identity
r2 <- GLMER(modelData = sites,responseVar = "Species_richness",
            fitFamily = "poisson",fixedStruct = "LandUse",
            randomStruct = "(1|SS)+(1|SSB)",REML = FALSE)

# Compare the AIC values of the models
print(AIC(r1$model,r2$model))
```

Clearly, including spatial block improves model fit, so we will include that random intercept in the models from now on.

Species richness models are often overdispersed. We can check whether that is the case as follows:

```{r, echo=TRUE}
GLMEROverdispersion(model = r2$model)
```

The residual deviance is much greater than the residual degrees of freedom. Consequently, the model is significantly over-dispersed. We can attempt to deal with this by including an observation-level random intercept (here site identity):

```{r, echo=TRUE}
r3 <- GLMER(modelData = sites,responseVar = "Species_richness",
            fitFamily = "poisson",fixedStruct = "LandUse",
            randomStruct = "(1|SS)+(1|SSB)+(1|SSBS)",REML = FALSE)

GLMEROverdispersion(model = r3$model)

AIC(r1$model,r2$model,r3$model)
```

This has removed the overdispersion. Also, the AIC of the model with site identity as a random intercept is much lower than the AIC of the other models. Therefore, we will proceed with this most-complex random-effects structure.

### Model Selection

We will now perform a very simple model selection exercise to test whether land use has a significant effect on species richness.

```{r, echo=TRUE}
m1 <- GLMERSelect(modelData = sites,responseVar = "Species_richness",
                  fitFamily = "poisson",fixedFactors = c("LandUse"),
                  randomStruct = "(1|SS)+(1|SSB)+(1|SSBS)",verbose = TRUE)
m1$stats
```
