--- 
title: "Modelling Land-use Impacts on Biodiversity - Getting Started with PREDICTS" 
author: "Tim Newbold" 
output: html_document 
---

This document will introduce the PREDICTS database, show you how to get the data, and how to start with some simple models of land-use impacts.

# Package dependencies
* predictsFunctions

The predictsFunctions and StatisticalModels packages can be installed from GitHub:

```{r, echo=TRUE, warning=FALSE, message=FALSE,results=FALSE}
library(remotes)
remotes::install_github("timnewbold/predicts-demo",subdir = "predictsFunctions")
remotes::install_github("timnewbold/StatisticalModels")
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, results=FALSE}
library(predictsFunctions)
library(StatisticalModels)
print(sessionInfo())
```

# The PREDICTS Database

## Getting the PREDICTS Database

You can get started quickly using the code below. If you intend to use the PREDICTS database in your work, please download it from the [Natural History Museum data portal](https://data.nhm.ac.uk/dataset/the-2016-release-of-the-predicts-database), to give credit for your use.

```{r, echo=TRUE}
# Download and import the PREDICTS data
myFile <- url("https://www.dropbox.com/s/pb1mdiel8o22186/database.rds?dl=1")
predicts <- readRDS(myFile)
print(head(predicts))
```

## Getting the Database Ready for Analysis

There are some standard steps to take before working with the PREDICTS database:
1. Correcting abundance values where sampling effort varies among the sites sampled in a single study.
2. Merging sites with the same coordinates.

```{r, echo=TRUE, message=FALSE, results=FALSE}
# Correct effort-sensitive abundance measures (assumes linear relationship between effort and recorded abundance)
predicts <- predictsFunctions::CorrectSamplingEffort(diversity = predicts)

# Merge sites that have the same coordinates (e.g. multiple traps on a single transect)
predicts <- predictsFunctions::MergeSites(diversity = predicts)
```

A lot of models of the PREDICTS database are of assemblage-level biodiversity measures. You can calculate these site-level measures using the *SiteMetrics* function.

```{r, echo=TRUE}
# Calculate site metrics of diversity - currently only species richness and total abundance
sites <- SiteMetrics(diversity = predicts,
                     extra.cols = c("Predominant_land_use",
                                    "SSB","SSBS","Biome"))
```

# Modelling the Impacts of Land Use

## Site-level biodiversity

We will start with some simple models of site-level biodiversity as a function only of land use.

First, we will rearrange the land-use classification a bit.

```{r, echo=TRUE}
sites$LandUse <- paste(sites$Predominant_land_use)
# Drop classification where land use could not be identified
sites$LandUse[(sites$LandUse=="Cannot decide")] <- NA
# Drop classification where the stage of recovery of secondary vegetation is unknown
sites$LandUse[(sites$LandUse=="Secondary vegetation (indeterminate age)")] <- NA
# Now make the variable a factor, and set the reference level to primary vegetation
sites$LandUse <- factor(sites$LandUse)
sites$LandUse <- relevel(sites$LandUse,ref="Primary vegetation")
```

We have to model the PREDICTS data using mixed-effects models, to account for the hierarchical structure of the database. We will start by investigating different possible random-effects structures for our models. All PREDICTS models need to have a random intercept of Study identity, since biodiversity measures are not comparable between studies.

We will compare here a model with a random intercept only of study, with one that contains a nested random intercept representing spatial blocks of sites within studies (to account for the spatial structuring of sites).

```{r, echo=TRUE}
# Fit a model with a random intercept only of study identity
r1 <- GLMER(modelData = sites,responseVar = "Species_richness",
            fitFamily = "poisson",fixedStruct = "LandUse",
            randomStruct = "(1|SS)",REML = FALSE)

# Try a model that also has a nested random intercept for spatial block identity
r2 <- GLMER(modelData = sites,responseVar = "Species_richness",
            fitFamily = "poisson",fixedStruct = "LandUse",
            randomStruct = "(1|SS)+(1|SSB)",REML = FALSE)

# Compare the AIC values of the models
print(AIC(r1$model,r2$model))
```

Clearly, including spatial block improves model fit.



